% note: make sure to replace all instances of procfs, debugfs, sysfs
% with their \texttt{} versions.
% presentation.tex
\documentclass{beamer}

% items enclosed in square brackets are optional; explanation below
\title{Unit tests for Linux kernel hackers}
\subtitle{How to avoid hours of painful debugging}
\author{Mihir Mehta}
\institute{
  Systems Core Group\\
  Samsung Research Institute - Noida\\[1ex]
  \texttt{mihir.mehta@samsung.com}
}
\date{February 3, 2014}

\begin{document}

%--- the titlepage frame -------------------------%
\begin{frame}[plain]
  \titlepage
\end{frame}

%--- the presentation begins here ----------------%
\begin{frame}{Overview}
  Overview of the material.
  \begin{itemize}

  \item What are unit tests?
  \item Why unit tests?
  %% \item Unit test design
  \item Accessing unit test results
  \item Summary

  \end{itemize}
\end{frame}

\begin{frame}{What are unit tests?}

\end{frame}

\begin{frame}{What are unit tests?}

  \begin{itemize}

  \item Ideally, a unit test involves calling a \textit{single} function
    and \textit{verifying} its return value against an
    \textit{expected} value.
  \item Running a test can have two outcomes.
    \begin{enumerate}
    \item Pass: Congrats! Your function works correctly on this test
      input!
    \item Fail: This again raises multiple outcomes:
      \begin{enumerate}
      \item Function body is wrong. (bad, but manageable if you know
        what you're doing.)
      \item Expected value is wrong (indicates you don't know what
        you're doing, i. e. very bad.)
      \end{enumerate}
    \end{enumerate}

  \end{itemize}

\end{frame}

\begin{frame}{Why unit tests?}

  \begin{itemize}

  \item Central component of test-driven design.
  \item Provide convenient way to verify our code.
  \item Save manual labour (i.e. less time at desk!)
  \item Crucial: help prevent regressions.

  \end{itemize}

\end{frame}

\begin{frame}{Accessing unit test results}

  \begin{itemize}
    \item In a userspace program, (i.e. what we're used to) it's
      simple to make a separate test executable and link the functions
      we are testing into this executable.
    \item This way, we maintain a logical separation of the test code
      from the codebase itself.
    \item We also avoid bloating our executables.
    \item Making a separate test executable is not possible within
      kernel code - why?
  \end{itemize}

\end{frame}

\begin{frame}{Accessing unit test results}

  \begin{itemize}
    \item Making a separate test executable is not possible within
      kernel code - why?
      \begin{itemize}
        \item Can't be a userspace program - userspace programs can't
          link to kernel code (privilege levels and other issues)
        \item Can't be a kernel - how are you going to run two kernels
          at once?
      \end{itemize}
    \item So, how do we implement unit tests for kernel code now?
  \end{itemize}

\end{frame}

\begin{frame}{Accessing unit test results}

  \begin{itemize}
    \item So, how do we implement unit tests for kernel code now?
      \begin{itemize}
      \item Kernel modules!
      \end{itemize}
    \item With modules, we can easily access kernel functions and test
      them.
    \item We can run all the tests at once, in the init function of
      the module, after setting a load priority for the module.
    \item If we use loadable modules (instead of static modules), we
      can avoid kernel bloat too - simply unload the module when done.
    \item One detail remains - how do we access the results of these
      tests in userspace?
  \end{itemize}

\end{frame}

\begin{frame}{Accessing unit test results}

  \begin{itemize}

  \item How do we access the results of our unit tests in userspace?
    \begin{itemize}
    \item Special kernel filesystems - sysfs and debugfs
    \end{itemize}

  \item It's actually quite common for kernel variables to need to be
    readable and possibly modifiable from  userspace. For this
    purpose, the Linux kernel provides certain special file systems.
    \begin{itemize}
    \item procfs - NOT to be used for debugging.
    \item sysfs - should not be used for debugging (but we do it anyway).
    \item debugfs - explicitly intended for debugging.
    \end{itemize}
  \item sysfs - used by many modules which need to communicate with
    user space; also abused for debugging; has strict restrictions on
    the types of exported values (only one numeric value per file).
  \item debugfs - used by many modules for debugging; has no
    restrictions on the types of exported values.
  \item Both of the above are sufficiently widely used to appear in
    nearly all distribution kernels (including ours).
  \item sysfs is generally mounted at \texttt{/sys}, while debugfs is
    generally mounted at \texttt{/sys/kernel/debug}.
  \end{itemize}

\end{frame}

\begin{frame}{Using debugfs for debugging}

  \begin{itemize}
  \item Every kernel module should store its debugging files in a
    separate directory in debugfs. Optionally, subdirectories can be
    used for better organisation.
  \item This function is used for creating a directory - \\
    \texttt{struct dentry *debugfs\_create\_dir(const char *name, struct
      dentry *parent);}\\
    If the value of \texttt{parent} is \texttt{NULL}, then the
    directory is created at the debugfs root.
  \item The \texttt{struct dentry *} pointer obtained from this
    function can now be used to create files in the directory, using
    this function - \\
    \texttt{struct dentry *debugfs\_create\_file(const char *name,
      umode\_t mode,
           struct dentry *parent, void *data,
           const struct file\_operations *fops);}
  \end{itemize}
  
\end{frame}

\begin{frame}{Using debugfs for debugging}

  \begin{itemize}

  \item Usually, though, the function
    \texttt{debugfs\_create\_file} is too bulky for our purposes -
    we'd rather not have to make a set of file operations.
  \item The debugfs module also provides a few convenience functions
    for common debugging use cases, to remove the need for making sets
    of file operations.
  \item In cases where we only want to export a numeric value, we can
    take advantage of these functions:
    \begin{itemize}

    \item \texttt{struct dentry *debugfs\_create\_u8(const char *name, 
      umode\_t mode,
           struct dentry *parent, u8 *value);}
    \item \texttt{struct dentry *debugfs\_create\_u16(const char *name,
      umode\_t mode,
          struct dentry *parent, u16 *value);}
    \item \texttt{struct dentry *debugfs\_create\_u32(const char *name,
      umode\_t mode,
          struct dentry *parent, u32 *value);}
    \item \texttt{struct dentry *debugfs\_create\_u64(const char *name,
      umode\_t mode,
          struct dentry *parent, u64 *value);}

    \end{itemize}

  \end{itemize}

\end{frame}

\begin{frame}{Acknowledgements}

  \begin{itemize}

  \item \texttt{doc/Documentation/filesystems/debugfs.txt} in the
    kernel source. (primary reference, authored by Jonathan Corbet.)

  \end{itemize}

\end{frame}

\end{document}
